---
title: "Covid-19 outcome prediction"
output:
  html_notebook: default
  pdf_document: default
---

This work is based on the data and Analysis of Li-Yan et al., 2019 published in Nature Machine Intelligence. 
https://www.nature.com/articles/s42256-020-0180-7#MOESM3

Blood analysis data from 485 patients infected by COVID-19 were collected, after the exclusion of pregnant women, breast-feeding woman, patient younger than 18 yo and patients for whom more than 20% of the data were missing. In the end,the data are divided in two sets: 375 individual in the first set (training set) and 110 in the second set (testing set). For each patient, up to 78 parameters were measured at several occasion between the admission and the release/death. 

Here we use the rapid glmnet method to produce a classification model based the last data point available for each parameter. We reasoned that using the last data point would be more predictive of the outcome. 

Based on the results of the glmnet model, 3 features selected to build an interpretable decision tree.

```{r load packages}
# load required libraries and set working directory

library(knitr)

library(caret) # for hyperparameters tunning
library(e1071) # dependency of caret
library(RANN) # for knn imputation

library(dplyr) 
library(tidyr)
library(ggplot2)
library(ggfortify) # enable auto plotting
library(rpart) # decision tree model
library(rattle) # plotting decision tree
library(glmnet) # glm model fitting

```


The identification of the patient (PATIENT_ID) is only displayed for the first row of each patient. 
```{r data loading}
#load data, perform clean-up
#propagate patient ID to all corresponding observations

cov_data = read.csv("time_series_375_prerpocess_en.csv",head= TRUE)
cov_data$RE_DATE = as.Date(cov_data$RE_DATE)
cov_data$Admission.time = as.Date(cov_data$Admission.time)
cov_data$Discharge.time = as.Date(cov_data$Discharge.time)

#fill-in patient ID
for (i in 1:length(cov_data$PATIENT_ID)) {
  cov_data$PATIENT_ID[i] = ifelse(is.na(cov_data$PATIENT_ID[i]) == TRUE,
                                  cov_data$PATIENT_ID[i - 1],
                                  cov_data$PATIENT_ID[i])
  
}

print(paste("number of patient: ", length(unique(cov_data$PATIENT_ID)))) # number of patients
head(cov_data)

```

A maximum of 81 parameters were collected for 375 patients
```{r data cleaning}
# select the LAST (most recent) available value of each parameter for each patient. The date of analysis is eliminated since the last value of different variable may have been taken at different dates (not every parameters are measured every day).
# eliminate parameters for which the value is unknown for more than 20% of patients
# turn the outcome into dead/alive instead of 1/0
# don't drop all columns containing NA values

time_col = c("RE_DATE","Admission.time","Discharge.time")

cov_mean <- cov_data %>%  group_by(PATIENT_ID) %>% 
                          arrange(RE_DATE) %>%
                          select(-time_col) %>%
                          summarise_all( ~ nth(.,max(which(!is.na(.))))) %>% 
                          select_if(funs(sum(is.na(.)) < 0.2 * length(unique(cov_data$PATIENT_ID))))  %>% 
                          mutate(outcome = as.factor(ifelse(outcome == 1, "dead", "alive"))) 

head(cov_mean)

print(paste("number of variables:", ncol(cov_mean)))

cov_mean_long = cov_mean[,-1] %>% pivot_longer(cols = colnames(cov_mean[,-1])[-3], names_to = "variables", values_to = "val")

cov_mean_long %>% head()


# create a boxplot for each variable in function of the outcome. The plot, generally to bit too large to be presented directly, is saved as a png.
multiplot = cov_mean_long %>% ggplot(aes(x = outcome, y = val, color = outcome)) +
                              geom_boxplot()+
                              facet_wrap(. ~ variables, ncol = 5, scales='free')+
                              labs(x = "outcome", y = "value")+
                              theme_classic()+
                              theme(legend.position = "none")

png("multiplot.png", res = 300, height = 10000, width = 5000)
multiplot
dev.off()

```
```{r PCA representation}

# Perform a PCA on x, with the column y as outcome. 
# Samples containing NA values are dropped (necessary to perform PCA)

prcom_NA = function(x,y){
  
  library(ggfortify)
  
  plot_colors <-  x %>% 
                  drop_na() %>%
                  select(y) %>%
                  unlist() %>%
                  as.factor()

  pr_plot =     x %>%  
                select(-c(y)) %>% 
                drop_na() %>% 
                prcomp(scale = TRUE, center = TRUE) %>%
                ggplot(aes(x = PC1, y =  PC2))+
                geom_point(aes(color = plot_colors), alpha = 0.7)+
                theme_classic()+
                theme(legend.title = element_blank())
  
  print(pr_plot)

} 


cov_mean %>% select(-PATIENT_ID) %>% prcom_NA("outcome")


```


```{r analysis of missing values}

library(naniar)
miss_var_summary(cov_mean)
miss_case_table(cov_mean)

gg_miss_var(cov_mean)
gg_miss_case(cov_mean)


# Are the samples with missing value very different from the rest ?
# calculate the relative difference for each parameters between the data with and without missing values

rel_diff <- cov_mean %>% bind_shadow(only_miss = TRUE) %>% 
            add_label_shadow() %>%
            group_by(any_missing) %>%
            select(!contains("_NA")) %>%
            mutate(outcome = ifelse(outcome == "alive",0,1)) %>%
            summarize_all(~ mean(., na.rm = TRUE)) %>%
            ungroup() %>%
            pivot_longer(-any_of("any_missing"), names_to = "variables", values_to = "val") 


rel_diff %>%  ggplot(aes(x = any_missing, y = log2(val), group = variables)) + 
              geom_line() + geom_point() + 
              theme_classic() + 
              labs(x = "")

library(ggrepel)


# for a histogram and density display of the relative difference between samples with missing variables and samples with no missing variables.
rel_diff %>%  pivot_wider(names_from = any_missing, values_from = val) %>% 
              dplyr::rename("Not_Missing"="Not Missing") %>%
              mutate(diff = (Missing-Not_Missing)/Missing) %>%
              filter(variables != "PATIENT_ID") %>%
              ggplot(aes(x = diff))+
              geom_histogram(fill = "lightgray", color = "white", bins = 50)+
              geom_density()+
              labs(x = "relative difference")+
              theme_minimal()


# for a violin and point display of the relative difference between samples with missing variables and samples with no missing variables.
rel_diff %>%  pivot_wider(names_from = any_missing, values_from = val) %>% 
              dplyr::rename("Not_Missing"="Not Missing") %>%
              mutate(diff = (Missing-Not_Missing)/Missing) %>%
              filter(variables != "PATIENT_ID") %>%
              ggplot(aes(x = "diff", y = diff))+
              geom_violin()+
              geom_point()+
              geom_text_repel(aes(label = ifelse(abs(diff) > 0.3, variables,"")), min.segment.length = 0, nudge_x = 0.2)+
              labs(x = " ", y = "relative difference sample w/o NA / samples with NA")+
              theme_classic()



```
As displayed on the graph, for the majority of the variables the differences between the samples where all variables are present and the samples where some are missing are smaller than 20%, suggesting that these samples are not fundamentally different. One can try to perform a knn or median imputation for the missing values.

One should however be careful if the model includes the concentration of procalicitonin, or the count of red blood cells or white blood cells.


```{r train and test samples preparation}

# Split the data in training and testing sets (70/30 %) 
set.seed(123)

index = as.vector(createDataPartition(cov_mean$PATIENT_ID, p = 0.7, list = FALSE))
cov_train = cov_mean[index,] %>% as.data.frame()
cov_test = cov_mean[-index,] %>% as.data.frame()

# missing values will be assigned with knn Imputation in the glmnet model. 

```


```{r glmnet fit}
# the original paper uses extreme gradient boosting for the task, here a rapid glmnet approach is presented.
# Alternatively random forest or xgbDART lead to similar results and can also be implemented in R.


# glmnet model fit to identify the most influential parameters in the prediction of the outcome
# 10 cross-validation with 5 repetitions
# predict the class, not the probability
# due to the difference of scale between parameters, the data are scaled and centered and the value of the missing data are imputed using knn if they have not been -1 padded before
# since missing data are imputed, one can pass the NA values to the function (na.action = na.pass)


fitControl = trainControl(method="repeatedcv", number=10, repeats = 5,
                          summaryFunction = twoClassSummary,
                          classProbs = TRUE,
                          verboseIter = F)


glmGrid = expand.grid(alpha = c(0,0.5,1),
                      lambda = seq(0.0001,0.1,length = 10))

imp_pred = list()
best_models = list()
glmfit_list = list()

#run multiple times for reproducibility (20)
print(Sys.time())

for (i in 1:20){
  
  print(paste("iteration",i))
  print(Sys.time())
  set.seed(i)

glmfit = train( outcome ~., 
                data = cov_train[,-1],
                method = "glmnet",
                metric = "ROC",
                trControl =fitControl,
                tuneGrid = glmGrid,
                preProcess = c("knnImpute", "center","scale"),
                standardize = TRUE,
                na.action = na.pass
                )
  
  imp_pred[[i]] = data.frame(var = rownames(varImp(glmfit)$importance), rel.inf = varImp(glmfit)$importance$Overall)
  best_models[[i]] = cbind(glmfit$bestTune,max(glmfit$results$ROC))
  glmfit_list[[i]] = glmfit
  
  print(paste("max accuracy", max(glmfit$results$ROC)))

}

plot(glmfit)


```

Display and select the most influential parameters
```{r display influential parameters}
# save the best hyper-parameters
best_a_l =  do.call(rbind, best_models) %>% 
              summarise_all(~median(.)) %>% 
              unlist()

head(best_a_l)

# select the most influential factors
# plot_inf <- do.call(rbind, imp_pred) %>%  group_by(var) %>% 
#                               summarise(rel_inf = mean(rel.inf)) %>% 
#                               top_n(10) %>%
#                               ggplot(aes(x = reorder(var,rel_inf), y = rel_inf))+
#                               geom_col(fill = "#00BFC4", alpha = 0.2)+
#                               labs(y = "relative influence", x = "parameters", title = "glmnet")+
#                               coord_flip()+
#                               theme_classic()
# 
# 
# print(plot_inf)

# display graph of the most influential parameters 
library(purrr)
do.call(rbind, imp_pred) %>%  as_tibble %>%
                              group_by(var) %>% 
                              nest() %>%
                              mutate(score = map_dbl(data, ~mean(.x$rel.inf))) %>%
                              mutate(sd = map_dbl(data, ~sd(.x$rel.inf))) %>%
                              select(-data) %>%
                              arrange(desc(score)) %>%
                              ungroup() %>%
                              slice(1:10) %>%
                              ggplot(aes(x = reorder(var,score), y = score)) + 
                              geom_point(fill = "#00BFC4", alpha = 0.2) +
                              coord_flip() +
                              labs(x = "parameter", y = "relative influence ± SD")+
                              geom_errorbar(aes(ymin = score-sd, ymax = score+sd), size = 0.1, width = 0.3)+
                              theme_classic()


```



Perform predictions on the test data 
```{r validate glmnet model on test data}

glmnet_pred_cm = list()

# Predict outcome with the best model of each iteration on test data
# The Models include knn imputation, NA values can be passed to the function

for(i in 1:20){

glmnet_pred = predict(glmfit_list[[i]], cov_test, na.action = na.pass)
glmnet_pred_cm[[i]] <- confusionMatrix(glmnet_pred, cov_test$outcome, positive = "dead")

}

# Display the average of the evaluation metrics
lapply(glmnet_pred_cm, "[[", "byClass") %>% do.call(rbind, .) %>% as_tibble %>% summarize_all(~mean(.))


```
mean metrics of predictions performed with the best model of each iteration on test data.



```{r features selection}

# select the optimal number of parameters acceptable for the decision tree model.
glmGrid = expand.grid(alpha = c(0, 0.5, 1),
                      lambda = seq(0.0001,0.1,length = 10))


# list the most influential parameters
top_10_param <- do.call(rbind, imp_pred) %>%  group_by(var) %>% 
                              summarise(rel_inf = mean(rel.inf)) %>% 
                              top_n(10) %>% arrange(-rel_inf)

nfeature = 1
best_depth = list()

for (nfeature in 1:5){
  
  best_models_min =list()
  
  # select parameters
  selected = c("outcome",top_10_param[1:nfeature,1]  %>% unlist() %>% as.character())
  
  # select values of the corresponding parameters
  cov_train_select <- cov_train[,selected]
  
  # known bug in the glmnet, it requires at least two columns of independent variables, add a column with constant value
  cov_train_select <- cov_train_select %>% mutate(ones = 1)

  print(paste("number of features:",nfeature))
  
      for (i in 1:10){
        
        print(paste("iteration:",i, "| number of features:",nfeature))
        print(Sys.time())
        set.seed(i)
      
      glmfit_minimal = train( outcome ~., 
                      data = cov_train_select,
                      method = "glmnet",
                      metric = "ROC",
                      trControl =fitControl,
                      tuneGrid = glmGrid,
                      preProcess = c("knnImpute","center","scale"),
                      standardize = TRUE,
                      na.action = na.pass
                      )
        
        best_models_min[[i]] = cbind(glmfit_minimal$bestTune,max(glmfit_minimal$results$ROC))
        
        print(paste("max ROC", max(glmfit_minimal$results$ROC)))
      
      }
  
  
  best_depth[[nfeature]] = mean(do.call(rbind,best_models_min)[,3])


}

best_d <-   do.call(rbind, best_depth) %>% as_tibble()

colnames(best_d) = "ROC"

best_d %>%  ggplot(aes(x = 1:nfeature, y = ROC)) + 
                geom_point() +
                geom_line()+
                theme_classic() +
                labs(x = "number of parameters", y = "ROC")


# select automatically an optimal number of features
best_d <- best_d %>% mutate(improvement = ROC - lag(ROC))
best_depth_val = which.max(best_d$improvement)

# or select manually the best balance between complexity and ROC, e.g.
best_depth_val = 3


```



```{r simple decision tree on the most influential parameters (rtree_model)}

# The optimal number of parameters has been selected
# Build a decision tree model based on these parameters

best_param <- plot_inf$data %>% arrange(desc(rel_inf)) %>%
              top_n(best_depth_val) %>%
              select(var) %>%
              unlist()%>%
              as.character() %>%
              append("outcome")

rControl <- trainControl(method="cv", number=10, 
                          summaryFunction = twoClassSummary,
                          classProbs = TRUE,
                          verboseIter = F)

rpart_model <- train(outcome ~., 
                cov_mean[,best_param],
                method = "rpart1SE",
                metric = "ROC",
                trControl =rControl,
                na.action = na.omit)

split1 = rpart_model$finalModel$splits[1,4]
split2 = rpart_model$finalModel$splits[2,4]

fancyRpartPlot(rpart_model$finalModel, sub = "rpart1SE")


```

```{r analysis of clinical evolution of the test 375 patients set}
## Select data by patient and by date (mean of the date collected each day if there are several measures)
# Select the most influential parameters (best_param)
# drop the row were the influential parameters are missing 
cov_data_all <- cov_data  %>% 
                group_by(PATIENT_ID, RE_DATE) %>% 
                summarise_all( ~ mean(., na.rm = TRUE)) %>% 
                select(c(RE_DATE,Discharge.time, best_param)) %>%
                drop_na() %>%
                mutate(outcome = (ifelse(outcome == 1, "dead", "alive")))  

cov_data_all$outcome = as.factor(cov_data_all$outcome)


# select observations of last date available for each patient
cov_data_lastday = cov_data_all %>% arrange(PATIENT_ID,RE_DATE) %>% 
                                    group_by(PATIENT_ID) %>%
                                    filter(RE_DATE == max(RE_DATE))

# plot patient trajectories as a line with dot representing the last measure
ggplot(cov_data_all, aes_string(x = best_param[1], y = best_param[2], group = "PATIENT_ID"))+
            geom_line(aes(color = as.factor(outcome)), alpha = 0.2)+
            theme_classic()+
            geom_hline(yintercept = split1, lty = 3, size = 0.5)+
            geom_segment(y = 0, yend = split1, x = split2, xend = split2,lty = 3, size = 0.5)+
            geom_point(data = cov_data_lastday, aes(color = as.factor(outcome)), alpha = 0.2)+
              facet_wrap(~outcome, nrow = 2)+
            labs(x = "hs-CRP (mg/L)", y = "LDH (U/L)", color = "outcome")
```
The graph above displays the evolution the 2 best parameters for each patients, with the dot representing the last available value. Apart for 2 patients, if the values of Lactate dehydrogenase and hsCRP are below 342 U/L and 42 mg/L respectively, the outcome was favorable. This suggest that this model could be useful to predict the clinical outcome well before the death or discharge.

```{r selecte all data with the selected parameters except the one corresponding to the last date}
# exclude observations of the last date for each patient
cov_data_firstday = cov_data_all %>% anti_join(cov_data_lastday, by = c("PATIENT_ID" = "PATIENT_ID", "RE_DATE" = "RE_DATE"))


pred_date = predict(rpart_model, cov_data_firstday, type = "raw" )
pred_matrix_test <- confusionMatrix(pred_date, cov_data_firstday$outcome, positive = "dead")
pred_matrix_test


```

```{r plot accuracy evolution}

# Inject the prediction to the original table and plot the average accuracy of the prediction in function of the time to outcome.

cov_data_firstday$pred = pred_date

cov_data_firstday = cov_data_firstday %>% mutate(result = (outcome == pred))
cov_data_firstday = cov_data_firstday %>% mutate(time_to_outcome = (Discharge.time - RE_DATE))

# plot the accuracy of the model depending on how close to the actual outcome together with the number of observations
cov_data_firstday %>% group_by(time_to_outcome) %>%
                  summarize(val = mean(result), val_sd = sd(result), val_c = n()) %>%
                  ggplot(aes(x = as.numeric(time_to_outcome), y = val_c))+
                  geom_col(fill = "#00BFC4", alpha = 0.3)+
                  geom_line(aes(y = val*25))+
                  theme_classic()+
                  scale_y_continuous("Number of observations", sec.axis = sec_axis(~ . * (1/25), name = "Accuracy ---", breaks = seq(0,1, by =0.2)))+
                  xlab("days to outcome")+
                  geom_hline(yintercept = pred_matrix_test$overall[1]*25, lty = 3)+
                  annotate("text", label = "average \n accuracy", x = 34, y = pred_matrix_test$overall[1]*25)

```
Starting 2 weeks before the outcome, the accuracy of the predictions made on the data from the 375 patients dataset gradually increases. (Since the the last data acquired before the outcome were used for the training, they were excluded from the testing set presented here and there is therefor no data available for testing the prediction on the day of the outcome)

```{r display prediction from (rtree_model)}



cov_data_firstday %>% ggplot(aes_string(x = best_param[1], y = best_param[2]))+
                    geom_point(aes(color = as.factor(result), shape = as.factor(outcome)), alpha = 0.7)+
                    geom_rug(aes(color = as.factor(result)), size = 0.2, alpha = 0.7)+
                    theme_classic()+
                    geom_hline(yintercept = split1, lty = 3, size = 0.5)+
                    geom_segment(y = 0, yend = split1, x = split2, xend = split2,lty = 3, size = 0.5)+
                    labs(color = "correct prediction")+
                    labs(shape = "outcome")+
                    labs(x = "hs-CRP (mg/L)", y = "LDH (U/L)")+
                    annotate("text", x = 250, y = 1500, label = paste("Accuracy all data except \n last day =",as.character(round(pred_matrix_test$overall[1],2))))


```
The incorrect predictions (in red) appear to be scattered.


```{r}
cov_data_firstday %>% ggplot(aes_string(x = best_param[1], y = best_param[2]))+
                    geom_point(aes(color = as.factor(outcome), shape = as.factor(result)), alpha = 0.7)+
                    geom_rug(aes(color = as.factor(outcome)), size = 0.2, alpha = 0.7)+
                    theme_classic()+
                    geom_hline(yintercept = split1, lty = 3, size = 0.5)+
                    geom_segment(y = 0, yend = split1, x = split2, xend = split2,lty = 3, size = 0.5)+
                    labs(color = "outcome")+
                    labs(shape = "correct prediction")+
                    labs(x = "hs-CRP (mg/L)", y = "LDH (U/L)")+
                    annotate("text", x = 250, y = 1500, label = paste("Accuracy all data except \n last day =",as.character(round(pred_matrix_test$overall[1],2))))
```


```{r}
# plotlyr

#install.packages("plotly")
library(plotly)

# 3D representation of the outcome of the patients from the training dataset in function of the selected parameters

fig3d <- plot_ly(data = cov_data_firstday, x = ~Lactate.dehydrogenase, y = ~Hypersensitive.c.reactive.protein, z = ~X...lymphocyte, color = ~outcome, colors = c('#F8776D', '#00BFC4'), opacity = 0.6, size = 0.3) 

fig3d <- fig3d %>% layout(
    title = "Selected features",
    scene = list(
      xaxis = list(title = "Lactate dehydrogenase (U/L)"),
      yaxis = list(title = "hs-CRP (mg/L)"),
      zaxis = list(title = "Lymphocytes (%)")
    ))

fig3d


```

The interpretable model can now be tested on an independent data set from 110 patients for every available dates.

```{r}

#test on new set of data
#load data, perform clean-up
#propagate patient ID to all corresponding observations

cov_new = read.csv("time_series_test_110_preprocess_en.csv",head= TRUE)
cov_new$RE_DATE = as.Date(cov_new$RE_DATE)
cov_new$Admission.time = as.Date(cov_new$Admission.time)
cov_new$Discharge.time = as.Date(cov_new$Discharge.time)

#fill-in patient ID
for (i in 1:length(cov_new$PATIENT_ID)) {
  cov_new$PATIENT_ID[i] = ifelse(is.na(cov_new$PATIENT_ID[i]) == TRUE,
                                  cov_new$PATIENT_ID[i - 1],
                                  cov_new$PATIENT_ID[i])
  
}

dim(cov_new)
length(unique(cov_new$PATIENT_ID))
head(cov_new)

cov_new_all =   cov_new  %>% 
                group_by(PATIENT_ID, RE_DATE) %>% 
                summarise_all( ~ mean(., na.rm = TRUE)) %>% 
                drop_na() %>%
                mutate(outcome = (ifelse(outcome == 1, "dead", "alive"))) 
               
cov_new_all$outcome =  as.factor(cov_new_all$outcome)

pred_date = predict(rpart_model, cov_new_all, type = "raw" )
pred_matrix_test <- confusionMatrix(pred_date, cov_new_all$outcome, positive = "dead")
pred_matrix_test

print(paste("Accuracy on test data from other days:",pred_matrix_test$overall[1]))

#inject predictions in the tibble
cov_new_all$pred = pred_date

cov_new_all = cov_new_all %>% mutate(result = (outcome == pred))
cov_new_all = cov_new_all %>% mutate(time_to_outcome = (Discharge.time - RE_DATE))

# plot the accuracy of the model depending on how close to the actual outcome together with the number of observations
cov_new_all %>%   filter(time_to_outcome >= 0) %>%
                  group_by(time_to_outcome) %>%
                  summarize(val = mean(result), val_sd = sd(result), val_c = n()) %>%
                  ggplot(aes(x = as.numeric(time_to_outcome), y = val_c))+
                  geom_col(fill = "#00BFC4", alpha = 0.3)+
                  geom_line(aes(y = val*25))+
                  theme_classic()+
                  scale_y_continuous("Number of observations", sec.axis = sec_axis(~ . * (1/25), name = "Accuracy ---", breaks = seq(0,1, by =0.2)))+
                  xlab("days to outcome")+
                  geom_hline(yintercept = pred_matrix_test$overall[1]*25, lty = 3)+
                  annotate("text", label = paste("average \n accuracy \n",round(pred_matrix_test$overall[1],4)), x = 32, y = pred_matrix_test$overall[1]*25)
                  
```
The model predicts with an overall 91 % accuracy the clinical outcome on an external data set.

```{r}

# plot patient trajectories form the external test dataset as a line with dot representing the last available measure
cov_new_lastday = cov_new_all %>% arrange(PATIENT_ID,RE_DATE) %>% 
                                    group_by(PATIENT_ID) %>%
                                    filter(RE_DATE == max(RE_DATE))

            ggplot(cov_new_all, aes_string(x = best_param[1], y = best_param[2], group = "PATIENT_ID"))+
            geom_line(aes(color = as.factor(outcome)), alpha = 0.2)+
            theme_classic()+
            geom_hline(yintercept = split1, lty = 3, size = 0.5)+
            geom_segment(y = 0, yend = split1, x = split2, xend = split2, lty = 3, size = 0.5)+
            geom_point(data = cov_new_lastday, aes(color = as.factor(outcome)), alpha = 0.2)+
            facet_wrap(~outcome, nrow = 2)+
            labs(y = "LDH U/L", x = "hs-CRP mg/L",color = "outcome")
            
```

The graph above displays the evolution the 2 best parameters for each patients, with the dot representing the last available value. On this new dataset, if the values of Lactate dehydrogenase and hsCRP are below 342 U/L and 42 mg/L respectively, the outcome was favorable. 



```{r}
# 3D representation of the outcome of the patients from the external test dataset in function of the selected parameters

fig3d_test <- plot_ly(data = cov_new_all, x = ~Lactate.dehydrogenase, y = ~Hypersensitive.c.reactive.protein, z = ~X...lymphocyte, color = ~outcome, colors = c('#F8776D', '#00BFC4'), opacity = 0.6, size = 0.3) 

fig3d_test <- fig3d_test %>% layout(
    title = "External dataset",
    scene = list(
      xaxis = list(title = "Lactate dehydrogenase (U/L)"),
      yaxis = list(title = "hs-CRP (mg/L)"),
      zaxis = list(title = "Lymphocytes (%)")
    ))

fig3d_test


```

  